### 2015-03-30 swmpr_manu.tex
finished proofing

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> install.packages('devtools')
> library(devtools)
> install_github('fawda123/SWMPr')
> library(SWMPr)
\end{verbatim}
\end{kframe}
\end{knitrout}

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for using each category of functions.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # retrieve metadata for all sites
> site_codes()
> 
> # retrieve metadata for a single site
> site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
>   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import a file from an independent server
> import_remote('acebbwq')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path/'
> 
> # import, do not include file extension
> import_local(path, 'apaebmet') 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> dat <- import_remote('apadbwq')
>  
> # view all attributes of dat
> attributes(dat)
>  
> # view a single attribute of dat
> attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view available methods for swmpr class
> methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # qaqc screen for a swmpr object, retain only '0'
> qaqc(dat)
> 
> # retain all data regardless of flag
> qaqc(dat, qaqc_keep = NULL)
> 
> # retain only '0' and '-1' flags
> qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view the number of observations in each QAQC flag
> qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> data(apaebmet)
> dat <- apaebmet
> 
> # select two parameters from dat
> subset(dat, select = c('rh', 'bp'))
> 
> # subset records greater than or equal to a date
> subset(dat, subset = '2013-01-01 0:00', operator = '>=')
> 
> # subset records within a date range
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
> 
> # subset records within a date range, select two parameters
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
>   select = c('atemp', 'totsorad'))
> 
> # remove rows/columns that do not contain data
> subset(dat, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import, qaqc removal
> data(apadbwq)
> dat <- qaqc(apadbwq)
> 
> # convert time series to two hour invervals
> # tolerance of +/- 30 minutes for matching existing data
> setstep(dat, timestep = 120, differ = 30)
> 
> # convert a nutrient time series to a continuous time series
> # then remove empty rows and columns
> data(apacpnut)
> dat_nut <- apacpnut
> dat_nut <- setstep(dat_nut, timestep = 60)
> subset(dat_nut, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nuts, wq, and met data as separate objects for the same reserve
> data(apacpnut)
> data(apacpwq)
> data(apaebmet)
> swmp1 <- apacpnut
> swmp2 <- apacpwq
> swmp3 <- apaebmet
> 
> # combine nuts and wq data by union
> comb(swmp1, swmp2, method = 'union')
> 
> # combine nuts and wq data by intersect
> comb(swmp1, swmp3, method = 'intersect')
> 
> # combine nuts, wq, and met data by nuts time series, two hour time step
> comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set periods of observation.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.  This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.  Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).  Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for \texttt{swmpr} objects to preserve the time series of the input data.

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggregate.swmpr} at the command line).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  Aggregation function is user-supplied but defaults to mean.\tabularnewline
\texttt{aggregate\_metab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw daily data.\tabularnewline
\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weatehr dataset using the open-water method.\tabularnewline
\texttt{decomp.swmpr}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \texttt{decompose}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\texttt{decomp\_cj.swmpr}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \texttt{decompTs} in the wq package.  Only monthly decomposition is possible.\tabularnewline
\texttt{hist.swmpr}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
\texttt{lines.swmpr}&Add lines to an existing \texttt{swmpr} plot.\tabularnewline
\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data.\tabularnewline
\texttt{plot.swmpr}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
\texttt{plot\_metab.swmpr}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.\tabularnewline
\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides can be specified, passed to \texttt{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # combine, qaqc, remove empty columns
> dat <- comb(swmp1, swmp2, method = 'union')
> dat <- qaqc(dat)
> swmpr_in <- subset(dat, rem_cols = T)
> 
> # get mean DO by quarters
> aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
> 
> # get mean DO by quarters, remove NA when calculating means
> fun_in <- function(x) mean(x, na.rm = T)
> aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
\end{verbatim}
\end{kframe}
\end{knitrout}

Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  This method is a simple wrapper to \texttt{filter}.  The \texttt{window} argument specifies the number of observations included in the moving average.  The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.  See \Cref{fig:smooth_ex} for the output from the code.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> data(apadbwq)
> swmp1 <- apadbwq
> 
> # qaqc and subset imported data
> dat <- qaqc(swmp1)
> dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
> 
> # filter
> test <- smoother(dat, window = 50, params = 'do_mgl')
> 
> # plot to see the difference
> plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
> lines(test, select = 'do_mgl', col = 'red', lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The \texttt{na.approx} function can be used to interpolate gaps in a \texttt{swmpr} object.  A required argument for the function is \texttt{maxgap} which defines the maximum gap size for interpolation.  See \Cref{fig:interp_ex} for the output from the following code.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apadbwq)
> swmp1 <- apadbwq
> 
> # qaqc and subset imported data
> dat <- qaqc(swmp1)
> dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
> 
> # interpolate, maxgap of 10 records
> test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
> 
> # interpolate maxgap of 30 records
> test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
> 
> # plot for comparison
> par(mfrow = c(3, 1))
> plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
> plot(do_mgl ~ datetimestamp, test, col = 'red', 
+   main = 'Interpolation - maximum gap of 10 records', type = 'l')
> lines(dat, select = 'do_mgl')
> plot(do_mgl ~ datetimestamp, test2, col = 'red', 
+   main = 'Interpolation - maximum gap of 30 records', type = 'l')
> lines(dat, select = 'do_mgl')
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.  The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.  By default, a moving average with a symmetric window is used to filter the seasonal component.  Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).

The \texttt{decompose} function requires a \texttt{ts} object with a specified frequency as input.  The \texttt{decomp} function converts the input \texttt{swmpr} vector to a \texttt{ts} object prior to \texttt{decompose}.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{`daily'} or \texttt{`seasonal'} can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.  Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.  \Cref{fig:decomp_ex1} is an example of the \texttt{decomp} function.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apadbwq)
> swmp1 <- apadbwq
> 
> # subset for daily decomposition
> dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
> 
> # decomposition and plot
> test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
> plot(test)
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

\caption[Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida]{Monthly aggregations of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data at Apalachicola Bay, Florida.\label{fig:metab_ex}}
\end{figure}

>>>>

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> install.packages('devtools')
> library(devtools)
> install_github('fawda123/SWMPr')
> library(SWMPr)
\end{verbatim}
\end{kframe}
\end{knitrout}

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for working with \ac{SWMP} data using each category of functions.

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # retrieve metadata for all sites
> site_codes()
> 
> # retrieve metadata for a single site
> site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
>   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import a file from an independent server
> import_remote('acebbwq')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path/'
> 
> # import, do not include file extension
> import_local(path, 'apaebmet') 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> dat <- import_remote('apadbwq')
>  
> # view all attributes of dat
> attributes(dat)
>  
> # view a single attribute of dat
> attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view available methods for swmpr class
> methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # qaqc screen for a swmpr object, retain only '0'
> qaqc(dat)
> 
> # retain all data regardless of flag
> qaqc(dat, qaqc_keep = NULL)
> 
> # retain only '0' and '-1' flags
> qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view the number of observations in each QAQC flag
> qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> data(apaebmet)
> dat <- apaebmet
> 
> # select two parameters from dat
> subset(dat, select = c('rh', 'bp'))
> 
> # subset records greater than or equal to a date
> subset(dat, subset = '2013-01-01 0:00', operator = '>=')
> 
> # subset records within a date range
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
> 
> # subset records within a date range, select two parameters
> subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
>   select = c('atemp', 'totsorad'))
> 
> # remove rows/columns that do not contain data
> subset(dat, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import, qaqc removal
> data(apadbwq)
> dat <- qaqc(apadbwq)
> 
> # convert time series to two hour invervals
> # tolerance of +/- 30 minutes for matching existing data
> setstep(dat, timestep = 120, differ = 30)
> 
> # convert a nutrient time series to a continuous time series
> # then remove empty rows and columns
> data(apacpnut)
> dat_nut <- apacpnut
> dat_nut <- setstep(dat_nut, timestep = 60)
> subset(dat_nut, rem_rows = T, rem_cols = T)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nut, wq, and met data as separate objects
> data(apacpnut)
> data(apacpwq)
> data(apaebmet)
> swmp1 <- apacpnut
> swmp2 <- apacpwq
> swmp3 <- apaebmet
> 
> # combine nut and wq data by union
> comb(swmp1, swmp2, method = 'union')
> 
> # combine nut and wq data by intersect
> comb(swmp1, swmp3, method = 'intersect')
> 
> # combine nut, wq, and met data by nut time series, two hour time step
> comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

The \texttt{aggregate} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data.

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggregate.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggregate.swmpr} at the command line).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{aggregate.swmpr}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\texttt{aggregate\_metab.swmpr}&Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
\texttt{ecometab.swmpr}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\texttt{decomp.swmpr}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\texttt{decomp\_cj.swmpr}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\texttt{hist.swmpr}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
\texttt{lines.swmpr}&Add lines to an existing plot created with \texttt{plot.swmpr}.\tabularnewline
\texttt{na.approx.swmpr}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\texttt{plot.swmpr}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
\texttt{plot\_metab.swmpr}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
\texttt{plot\_summary.swmpr}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\texttt{smoother.swmpr}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apacpwq)
> dat <- qaqc(apacpwq)
> 
> # get mean DO by months
> aggregate(dat, 'months', params = c('do_mgl'))
\end{verbatim}
\end{kframe}
\end{knitrout}

Time series can be smoothed to better characterize a signal independent of noise (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size with arguments passed to the \texttt{filter} function from the base installation.  The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \texttt{sides} argument specifies how the average is calculated for each observation.  Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data, qaqc and subset
> data(apadbwq)
> dat <- qaqc(apadbwq)
> dat <- subset(dat, select = 'do_mgl', 
+   subset = c('2012-07-09 00:00', '2012-07-24 00:00')
+   )
> 
> # smooth
> dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
> 
> # plot raw and smoothed
> plot(dat, type = 'l')
> lines(dat_smooth, select = 'do_mgl', col = 'red', lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data, qaqc and subset
> data(apadbwq)
> dat <- qaqc(apadbwq)
> dat <- subset(dat, select = 'do_mgl', 
+   subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
> 
> # interpolate, maxgap of 10 records
> fill1 <- na.approx(dat, params = 'do_mgl', maxgap = 10)
> 
> # interpolate maxgap of 30 records
> fill2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
> 
> # plot for comparison
> par(mfrow = c(3, 1))
> plot(dat, main = 'Raw')
> plot(fill1, col = 'red', main = 'Interpolation - maximum gap of 10 records')
> lines(dat)
> plot(fill2, col = 'red', main = 'Interpolation - maximum gap of 30 records')
> lines(dat)
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

The disaggregation of time series into additive of multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation (e.g., 1 if the the first observation is at sunrise for a diurnal cycle).\footnote{The value is passed to the \texttt{ts} function, \url{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}}  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apadbwq)
> swmp1 <- apadbwq
> 
> # subset for daily decomposition
> dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
> 
> # decomposition and plot
> test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
> plot(test)
\end{verbatim}
\end{kframe}\begin{figure}[!ht]

@@

\caption[Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida]{Monthly means (95\% confidence) of ecosystem metabolism estimates (net ecosystem metabolism, gross production, and total respiration) for combined water quality and weather data for two years at Apalachicola Bay, Florida.\label{fig:metab_ex}}
\end{figure}
